---
title: "01 – Shift-Representability: Core Tests"
description: "Demonstrating the numerical toolkit: single-family checks and batch parameter sweeps."
---

```{r}
#| label: init
#| message: false
library(distributional)
library(ggplot2)
library(purrr)
library(dplyr)
library(tidyr)

source("R/utils.R")
source("R/shift_tests.R")

dir.create("output", showWarnings = FALSE)
```

## Background

A finite family of random variables $\{X_1, \ldots, X_k\}$ is
**shift-representable** if there exists a strictly monotone $g$ such that the
transformed family $\{g(X_n)\}$ is a location (shift) family.  Two *necessary*
conditions (Theorem 2) are:

1. **Commutativity** – the pairwise quantile-matching transport maps
   $\psi_n = F_1^{-1} \circ F_n$ must commute under composition.
2. **Non-crossing** – no two transport maps may intersect, and no transport map
   may have a fixed point (i.e., $F_n$ may not cross $F_1$).

---

## Example 1 – Gamma family (shift-representable)

$X_n \sim \text{Gamma}(\alpha, \beta_n)$ with **common shape** $\alpha$ and
varying rate $\beta_n$.  A log transformation maps this family to an Extreme
Value location family, so it is shift-representable and both conditions should
pass.

```{r}
#| label: gamma-family
gamma_family <- distribution_family(
  X1 = dist_gamma(shape = 5, rate = 3),
  X2 = dist_gamma(shape = 5, rate = 2),
  X3 = dist_gamma(shape = 5, rate = 1.2)
)
gamma_family
```

```{r}
#| label: gamma-transport-maps
x_grid_gamma <- seq(0.01, 10, length.out = 400)

gamma_maps <- eval_transport_maps(gamma_family, x_grid_gamma)

ggplot(gamma_maps, aes(x = x, y = psi_x, colour = map_id)) +
  geom_line(linewidth = 0.9) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    title    = "Transport maps – Gamma family (common shape)",
    x        = "x  (reference scale)",
    y        = expression(psi[n](x)),
    colour   = "Map"
  ) +
  theme_bw()
```

```{r}
#| label: gamma-check
shift_representability_check(gamma_family, x_grid_gamma)
```

---

## Example 2 – Unequal-variance normals (not shift-representable)

$N \sim \mathcal{N}(0, 1)$, $S_i \sim \mathcal{N}(\mu_i, \sigma_i^2)$ with
$\sigma_i \neq 1$.  The transport maps are **affine dilations** – they intersect
the identity and each other, so commutativity fails.

```{r}
#| label: uvsd-family
means <- c(0, 1, 2)
sds <- c(1, 1.3, 1.3)
uvsd_family <- dist_normal(means, sds)
attr(uvsd_family, "reference") <- 1
```

```{r}
#| label: uvsd-transport-maps
x_grid_norm <- seq(-4, 6, length.out = 400)

uvsd_maps <- eval_transport_maps(uvsd_family, x_grid_norm)

ggplot(uvsd_maps, aes(x = x, y = psi_x, colour = map_id)) +
  geom_line(linewidth = 0.9) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    title  = "Transport maps – Unequal-variance normals",
    x      = "x  (reference scale)",
    y      = expression(psi[n](x)),
    colour = "Map"
  ) +
  theme_bw()
```

```{r}
#| label: uvsd-check
uvsd_check <- shift_representability_check(uvsd_family, x_grid_norm)
uvsd_check
```

### Commutativity deviation

```{r}
#| label: uvsd-comm-plot
ggplot(uvsd_check$commutativity$pointwise,
       aes(x = x, y = deviation, colour = interaction(map_n, map_m))) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    title  = "Commutativity deviation: ψ_n(ψ_m(x)) − ψ_m(ψ_n(x))",
    x      = "x",
    y      = "Deviation",
    colour = "Pair (n.m)"
  ) +
  theme_bw()
```

---

## Example 3 – UVSD with common crossing point (weakly shift-representable)

When $\sigma_i = 1 + a \mu_i$ for a common $a$, all distributions share the
same crossing point $x^* = -1/a$, commutativity is restored, and the family is
shift-representable on the one-sided domain $x > x^*$.

```{r}
#| label: uvsd-common-crossing
a <- 0.25   # σ = 1 + 0.25 * μ
mu_vals <- c(0, 1, 2, 3)
sd_vals <- 1 + a * mu_vals
x_star  <- -1 / a   # common crossing: x* = -4

uvsd_cc_dists <- dist_normal(mu_vals, sd_vals)
uvsd_cc_family <- do.call(distribution_family, uvsd_cc_dists)

# Restrict domain to x > x* (one-sided, shift-representable region)
x_grid_cc <- seq(x_star + 0.1, 6, length.out = 400)

shift_representability_check(uvsd_cc_family, x_grid_cc)

```

---

## Example 4 – Batch sweep: when does a 3-Gamma family fail?

Commutativity only becomes a binding constraint with **three or more**
distributions (any pair is always shift-representable by Abel's theorem).  Fix
$X_1 \sim \text{Gamma}(5, 1)$ as the reference and sweep over
$(\alpha_2, \alpha_3)$ — the shapes of $X_2$ and $X_3$.

When $\alpha_2 = \alpha_3 = 5$ (all shapes equal) the family is
shift-representable.  Any mismatch breaks commutativity.

```{r}
#| label: gamma-sweep
ref_shape <- 5
shapes     <- c(2, 3, 4, 5, 6, 8)
x_grid_sw  <- seq(0.05, 20, length.out = 1000)

sweep_results <- expand_grid(shape2 = shapes, shape3 = shapes) |>
    mutate(
      family = map2(shape2, shape3, \(s2, s3) {
        distribution_family(
          X1 = dist_gamma(shape = ref_shape, rate = 1),
          X2 = dist_gamma(shape = s2,        rate = 1),
          X3 = dist_gamma(shape = s3,        rate = 1)
        )
      }),
      check  = map(family, shift_representability_check, x_grid = x_grid_sw),
      result = map(check, summarise_check)
    ) |> 
    unnest(result)

ggplot(sweep_results, aes(x = factor(shape2), y = factor(shape3),
                           fill = log10(max_commutativity_dev + 1e-12))) +
  geom_tile() +
  geom_text(aes(label = ifelse(passes_necessary, "PASS", "FAIL")), size = 3) +
  scale_fill_viridis_c(name = "log₁₀(max |dev|)", option = "plasma") +
  labs(
    title    = "Commutativity violations – 3-Gamma family (X_1, shape = 5, rate = 1)",
    subtitle = "Shape of X_2 vs. shape of X_3",
    x        = "shape₂",
    y        = "shape₃"
  ) +
  theme_bw()
```

Some are weird

```{r}
sweep_results$family[[4]]
```
