---
title: "Transport Invariance: Gamma Distribution Test"
format: html
---

This notebook tests the transport-invariance identification algorithm using
gamma distributions. Gamma CDFs with the same shape parameter are related by
pure scaling of the rate, which makes them transport-invariant under the
probability-scale transport maps. The algorithm should recover this: given two
gamma exemplars with shared shape but different rates, a candidate gamma(shape,
rate) should be found with the same shape as the exemplars and near-zero
objective.

```{r}
#| label: init
source("R/transport-invariance.R")
```

## Define exemplar gamma distributions

We pick two gamma distributions with the same shape but different rates as our
exemplars. The reference is exemplar 1.

```{r}
#| label: define-exemplars
shape_true <- 3
rate_1 <- 1
rate_2 <- 2.5

exemplar_cdfs <- list(
  function(x) pgamma(x, shape = shape_true, rate = rate_1),
  function(x) pgamma(x, shape = shape_true, rate = rate_2)
)

Q_1 <- function(u) qgamma(u, shape = shape_true, rate = rate_1)
```

## Visualize the exemplar CDFs and their transport map

```{r}
#| label: plot-exemplars
x_grid <- seq(0.01, 10, length.out = 500)
plot(x_grid, exemplar_cdfs[[1]](x_grid), type = "l", col = "blue", lwd = 2,
     xlab = "x", ylab = "F(x)", main = "Exemplar Gamma CDFs")
lines(x_grid, exemplar_cdfs[[2]](x_grid), col = "red", lwd = 2)
legend("bottomright", legend = c(
  sprintf("Gamma(%g, %g)", shape_true, rate_1),
  sprintf("Gamma(%g, %g)", shape_true, rate_2)
), col = c("blue", "red"), lwd = 2)
```

```{r}
#| label: plot-transport
u_grid <- seq(0.01, 0.99, length.out = 500)
phi_2 <- exemplar_cdfs[[2]](Q_1(u_grid))

plot(u_grid, phi_2, type = "l", col = "darkgreen", lwd = 2,
     xlab = "u", ylab = expression(phi[2](u)),
     main = "Probability-scale transport map")
abline(0, 1, lty = 2, col = "gray")
```

## Search for the transport-invariant gamma member

The candidate family is `Gamma(shape, rate)` with both parameters free. The
algorithm should find that the shape must match the exemplars' shape, while the
rate is unconstrained (any rate yields a valid invariant member).

```{r}
#| label: define-candidate-factory
candidate_cdf_factory <- function(theta) {
  s <- theta[1]
  r <- theta[2]
  function(x) pgamma(x, shape = s, rate = r)
}
```

```{r}
#| label: run-optimization
result <- find_invariant_member(
  exemplar_cdfs = exemplar_cdfs,
  Q_1 = Q_1,
  candidate_cdf_factory = candidate_cdf_factory,
  theta_lower = c(1, 1),
  theta_upper = c(20, 20),
  n_starts = 30,
  m = 2000,
  eps = 1e-3
)
```

## Results

```{r}
#| label: print-results
cat("Best theta (shape, rate):", result$theta_star, "\n")
cat("Objective value J(theta*):", result$objective, "\n")
cat("True exemplar shape:", shape_true, "\n")
cat("Recovered shape:", round(result$theta_star[1], 4), "\n")
cat("Recovered rate:", round(result$theta_star[2], 4), "\n")
cat("\nAll objectives across starts:\n")
print(summary(result$all_objectives))
```

## Diagnostics

### Objective landscape over shape (fixing rate at recovered value)

We expect a clear minimum at the true shape.

```{r}
#| label: objective-profile-shape
pre <- result$precomputed

shape_grid <- seq(0.5, 8, length.out = 200)
obj_shape <- vapply(shape_grid, function(s) {
  compute_objective(
    theta = c(s, result$theta_star[2]),
    u_mesh = pre$u_mesh, x_mesh = pre$x_mesh,
    phi_values = pre$phi_values,
    exemplar_cdfs = exemplar_cdfs,
    Q_1 = Q_1,
    candidate_cdf_factory = candidate_cdf_factory
  )
}, numeric(1))

plot(shape_grid, obj_shape, type = "l", lwd = 2,
     xlab = "shape", ylab = "J(shape, rate*)",
     main = "Objective profile over shape (rate fixed)")
abline(v = shape_true, lty = 2, col = "red")
legend("topright", "true shape", lty = 2, col = "red")
```

### Objective landscape over rate (fixing shape at recovered value)

We expect the objective to be near zero for all rate values when shape matches.

```{r}
#| label: objective-profile-rate
rate_grid <- seq(0.1, 10, length.out = 200)
obj_rate <- vapply(rate_grid, function(r) {
  compute_objective(
    theta = c(result$theta_star[1], r),
    u_mesh = pre$u_mesh, x_mesh = pre$x_mesh,
    phi_values = pre$phi_values,
    exemplar_cdfs = exemplar_cdfs,
    Q_1 = Q_1,
    candidate_cdf_factory = candidate_cdf_factory
  )
}, numeric(1))

plot(rate_grid, obj_rate, type = "l", lwd = 2,
     xlab = "rate", ylab = "J(shape*, rate)",
     main = "Objective profile over rate (shape fixed at recovered)")
```

### Convergence across multi-starts

```{r}
#| label: convergence-summary
theta_all <- do.call(rbind, lapply(result$all_results, function(r) r$par))
colnames(theta_all) <- c("shape", "rate")

cat("Recovered shape across all starts:\n")
print(summary(theta_all[, "shape"]))
cat("\nRecovered rate across all starts:\n")
print(summary(theta_all[, "rate"]))

cat("\nSD of recovered shapes:", sd(theta_all[, "shape"]), "\n")
cat("SD of recovered rates:", sd(theta_all[, "rate"]), "\n")
cat("\nNote: rate should vary (not unique), shape should be consistent.\n")
```

### Scatter of converged parameters

```{r}
#| label: scatter-starts
plot(theta_all[, "shape"], theta_all[, "rate"],
     pch = 19, col = adjustcolor("steelblue", 0.7),
     xlab = "shape", ylab = "rate",
     main = "Converged parameters across multi-starts")
abline(v = shape_true, lty = 2, col = "red")
legend("topright", "true shape", lty = 2, col = "red")
```
