---
title: "Using `plot_commutator` to Diagnose Transport Invariance"
format: html
---

`plot_commutator` visualizes the commutator residual
$H_\theta \circ \phi_i(u) - \phi_i \circ H_\theta(u)$ as a function of $u$
for each non-reference exemplar. When a candidate distribution is
transport-invariant with respect to the exemplars, the residual is identically
zero. Deviations from zero reveal which candidates violate invariance and how.

```{r}
#| label: init
source("R/transport-invariance.R")
```

## Setup: gamma exemplars and candidate family

We use two Gamma distributions with the same shape but different rates. Members
of the family with matching shape are transport-invariant; members with
mismatched shape are not.

```{r}
#| label: start
exemplar_cdfs <- list(
  function(x) pgamma(x, shape = 3, rate = 1),
  function(x) pgamma(x, shape = 3, rate = 2.5)
)
Q_1 <- function(u) qgamma(u, shape = 3, rate = 1)

candidate_cdf_factory <- function(theta) {
  function(x) pgamma(x, shape = theta[1], rate = theta[2])
}
```

## Correct shape: residuals near zero

When the candidate has the same shape as the exemplars, it is
transport-invariant regardless of the rate. The commutator residual should be
flat at zero.

```{r}
#| label: correct-shape
plot_commutator(c(3, 1.5), exemplar_cdfs, Q_1, candidate_cdf_factory,
  main = "Correct shape (transport-invariant)")
```

Changing the rate while keeping the correct shape still yields zero residuals:

```{r}
#| label: correct-shape-other-rate
plot_commutator(c(3, 5), exemplar_cdfs, Q_1, candidate_cdf_factory,
  main = "Correct shape, different rate (still invariant)")
```

## Wrong shape: non-zero residuals

When the candidate's shape parameter differs from the exemplars', the
commutator is non-zero, indicating a violation of transport invariance.

```{r}
#| label: wrong-shape
plot_commutator(c(5, 1.5), exemplar_cdfs, Q_1, candidate_cdf_factory,
  main = "Wrong shape = 5 (not transport-invariant)")
```

A smaller deviation in shape produces smaller but still non-zero residuals:

```{r}
#| label: slight-mismatch
plot_commutator(c(3.5, 1.5), exemplar_cdfs, Q_1, candidate_cdf_factory,
  main = "Slight shape mismatch = 3.5")
```

## Comparing residual magnitude across shapes

We can sweep over a range of shape values and plot the peak absolute residual
to see how the violation grows with shape mismatch.

```{r}
#| label: residual-sweep
shape_grid <- seq(1, 8, length.out = 50)
pre <- precompute(exemplar_cdfs, Q_1, m = 1000, eps = 1e-3)

max_residuals <- vapply(shape_grid, function(s) {
  res <- compute_commutator(
    theta = c(s, 1.5),
    u_mesh = pre$u_mesh,
    x_mesh = pre$x_mesh,
    phi_values = pre$phi_values,
    exemplar_cdfs = exemplar_cdfs,
    Q_1 = Q_1,
    candidate_cdf_factory = candidate_cdf_factory
  )
  max(abs(unlist(res)))
}, numeric(1))

plot(shape_grid, max_residuals, type = "l", lwd = 2,
  xlab = "candidate shape", ylab = "max |commutator residual|",
  main = "Residual magnitude vs. candidate shape")
abline(v = 3, lty = 2, col = "red")
legend("topright", "true shape = 3", lty = 2, col = "red")
```

## Multiple exemplars

`plot_commutator` handles any number of non-reference exemplars. Each gets its
own curve.

```{r}
#| label: three-exemplars
exemplar_cdfs_3 <- list(
  function(x) pgamma(x, shape = 3, rate = 1),
  function(x) pgamma(x, shape = 3, rate = 2),
  function(x) pgamma(x, shape = 3, rate = 4)
)
Q_1_3 <- function(u) qgamma(u, shape = 3, rate = 1)

plot_commutator(c(3, 1), exemplar_cdfs_3, Q_1_3, candidate_cdf_factory,
  main = "Three exemplars, correct shape")
```

```{r}
#| label: three-exemplars-wrong
plot_commutator(c(5, 1), exemplar_cdfs_3, Q_1_3, candidate_cdf_factory,
  main = "Three exemplars, wrong shape")
```


### Tests


```{r}
#| label: three-exemplars2
exemplar_cdfs_3 <- list(
  function(x) pgamma(x, shape = 3, rate = 1),
  function(x) pgamma(x, shape = 4, rate = 1.15)
)
Q_1_3 <- function(u) qgamma(u, shape = 3, rate = 1)

plot_commutator(c(6,1.43), exemplar_cdfs_3, Q_1_3, candidate_cdf_factory,
  main = "Three exemplars, correct shape")
```

## Objective heatmap

`plot_objective_heatmap` evaluates the objective over a 2D parameter grid and
displays a log-scaled heatmap with contour lines. The valley at shape = 3
(spanning all rates) is clearly visible.

```{r}
#| label: heatmap

exemplar_cdfs <- list(
  function(x) pgamma(x, 1, 1),
  function(x) pgamma(x, 2, 0.95)
)
candidate_cdf_factory <- function(theta) {
  function(x) pgamma(x, theta[1], theta[2])
}
Q_1 <- function(x) qgamma(x, 1, 1)
plot_objective_heatmap(
  theta1_grid = seq(0.01, 10, length.out = 200),
  theta2_grid = seq(0.4, 2, length.out = 200),
  exemplar_cdfs = exemplar_cdfs,
  Q_1 = Q_1,
  candidate_cdf_factory = candidate_cdf_factory,
  overlay = function() abline(v = 2.5, col = "black")
)
```

```{r}
exemplar_cdfs <- list(
  function(x) pgamma(x, 1, 1),
  function(x) pgamma(x, 1.1, 0.5)
)
candidate_cdf_factory <- function(theta) {
  function(x) pgamma(x, theta[1], theta[2])
}
Q_1 <- function(x) qgamma(x, 1, 1)
result <- find_invariant_member(
  exemplar_cdfs = exemplar_cdfs,
  Q_1 = Q_1,
  candidate_cdf_factory = candidate_cdf_factory,
  theta_lower = c(1.15,0.1),
  theta_upper = c(40,20),
  n_starts = 5,
  m = 2000,
  eps = 1e-4
)
str(result, digits = 10)
print(result$theta_star, digits = 15)
```

```{r}
exemplar_cdfs_3 <- list(
  function(x) pgamma(x, shape = 1, rate = 1),
  function(x) pgamma(x, shape = 1.1, rate = 0.9)
)
Q_1_3 <- function(u) qgamma(u, shape = 1, rate = 1)

plot_commutator(c(10.2,0.1), exemplar_cdfs_3, Q_1_3, candidate_cdf_factory,
  main = "Three exemplars, correct shape")
```


